\section{Introduction}

This paper is a focused companion to the public order-four magic-square work.
Its object is narrower: the certified endpoint-24 boundary atlas in
affine-defect coordinates.  The earlier paper isolates \(24\) as a structured
bounded meet.  Here we examine the finite boundary that appears when the
order-four endpoint-24 atlas is organized by affine error over \(\F_2^4\).

The central split is
\[
  236 = 144 + 32 + 60,
\]
where the three terms are the exact affine core, the selected-affine B2
boundary, and the selected-non-affine B3 boundary.  The verified free
\(\TK\cong V_4\) action gives the quotient split
\[
  59 = 36 + 8 + 15.
\]

The exact affine core has \(144\) records.  The selected-affine boundary B2 has
\(32\) records.  The selected-non-affine boundary B3 has \(60\) records and
carries a support-action matrix with \(15\) realized and \(49\) absent
signature/profile cells.  B1 is included as a scoped companion statement: the
endpoint \(23/27\) signatures are absent from the affine selected-plane layer by
the same four-point xor criterion, without claiming a full endpoint \(23/27\)
classification.

\paragraph{Contribution.}
The paper has three layers.
\begin{enumerate}[leftmargin=*]
\item It states the endpoint-24 affine-defect boundary theorem in a compact
  certified-atlas form.
\item It separates the structural part of the proof from the finite matrix
  part.  The degree/defect dichotomy and the error-fiber parity bridge are
  proved as first-principles finite theorems; the B3 support-action matrix is
  a finite transport certificate.
\item It collapses the remaining B3 count debt from ``why \(15\) orbits'' to
  the sharper finite kernel
  \[
    \delta=0,4,6\mapsto 36,16,7
  \]
  and the B3 row-degree pattern \(1,1,1,2,2,1,4,3\).
\end{enumerate}

\paragraph{Claim discipline.}
Replay-backed statements are labelled as Certified Propositions and point to
the source map in \texttt{paper/SOURCE\_MAP.md}.  The manuscript does not claim
a non-enumerative derivation of the B3 count \(60\), a symbolic derivation of
the endpoint kernel \(36,16,7\), a full endpoint \(23/27\) classification, or a
full classification of non-affine order-four magic-square behavior.